\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-08-25 paper_OJA.tex
MOPED (begin an introduction)

\subsection{Instantaneous Fisher Forecasts}
In this section, we illustrate with a purely fictional case study, the computation of the Fisher matrix by two methods thanks to the \textit{autodiff} ability of JAX. For the detailed implementation, the reader is invited to have a look at the following companion notebook \nblink{Simple-Fisher}. In this use-case we use four tracers redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels. Then, the $10$ angular power spectra $C_\ell^{i,j}$ ($i,j:1,\dots,4$) are computed for $50$  logarithmically spaced angular moments between $10$ and $1000$ using Equation \ref{eq:Cell_limber}. The Gaussian covariance matrix is computed simultaneously. A dataset is obtained from the computation of the $C_\ell^{i,j}$ with a fiducial cosmology. Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation considering a constant covariance matrix ($\theta$ stands for the set of cosmological parameters). 
%\begin{minted}[fontsize=\footnotesize]{python}
\begin{lstlisting}[language=iPython]
@jax.jit
def likelihood(p):
    # Create a new cosmology at these parameters
    cosmo = jc.Planck15(Omega_c=p[0], sigma8=p[1])
    # Compute mean and covariance of angular Cls
    mu, C = jc.angular_cl.gaussian_cl_covariance_and_mean(cosmo, ell, tracers, sparse=True)
    # Return likelihood value assuming constant covariance, so we stop the gradient
    # at the level of the precision matrix, and we will not include the logdet term
    # in the likelihood
    P = jc.sparse.inv(jax.lax.stop_gradient(C))
    r = data - mu
    return -0.5 * r.T @ jc.sparse.sparse_dot_vec(P, r)
\end{lstlisting}
%\end{minted}
The \texttt{jc.sparse} functions are implementation of block matrix computations: sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_diag)$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.  The \texttt{jax.jit} decorator is there to build a compiled version of the function at the first use. The goal is to obtain the 1-sigma contour forecast of the two parameters ($\Omega_c, \sigma_8$). To do so, we use the Fisher matrix computation. One way is to use the Hessian of the log-likelihood as follows
\begin{equation}
F_{\alpha\beta} = - \frac{\partial^2\mathcal{L}(\theta)}{\partial \theta_\alpha \partial \theta_\beta}
\qquad (\theta_\alpha=\Omega_c, \theta_\beta=\sigma_8)
\label{eq:fisher_way1}
\end{equation}
which is mostly  1-line JAX code
\begin{lstlisting}[language=iPython]
hessian_loglik = jax.jit(jax.hessian(likelihood))
F = - hessian_loglik(params)
\end{lstlisting}
An other way uses the computation of the Fisher matrix when the likelihood is a Gaussian distribution as ii is the case above. First, we define a function that computes the mean  $\mu(\ell; \theta)$, and it follows that the Fisher matrix element reads
\begin{equation}
    F_{\alpha\beta} = \sum_\ell \frac{\partial \mu(\ell)}{\partial \theta_\alpha}^T C^{-1}(\ell)\frac{\partial \mu(\ell)}{\partial \theta_\beta}
    \label{eq:fisher_way2}
\end{equation}
where $C^{-1}(\ell)$ is the covariance matrix computed with the fiducial cosmology. In this case one needs also just 1-line of JAX code 
\begin{lstlisting}[language=iPython]
F = jc.sparse.dot(dmu.T, jc.sparse.inv(cov), dmu)
\end{lstlisting}

@@

\section{Application: Classical Hamiltonian Monte Carlo}
\label{sec:chmc}
%
In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo. We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI). All  the methods have been implemented using the \numpyro\ probabilistic programming language (PPL). 
%
%
%
\subsection{Vanilla HMC}
%
% Show Joe's vanilla HMC results against Cobaya
Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
and developed extensively since.  See \citet{betancourt} for a full review; we describe
very basic features here.

>>>>

\subsection{Instantaneous Fisher Forecasts}
In this section, we illustrate with a purely fictional case study, the computation of the Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} by two methods thanks to the \textit{autodiff} ability of JAX. For the detailed implementation, the reader is invited to have a look at the following companion notebook \nblink{Simple-Fisher}. In this use-case we use four tracers redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels. Then, the $10$ angular power spectra $C_\ell^{p,q}$ ($p,q:1,\dots,4$) are computed for $50$  logarithmically spaced angular moments between $10$ and $1000$ using Equation \ref{eq:Cell_limber}. The Gaussian covariance matrix is computed simultaneously. A dataset is obtained from the computation of the $C_\ell^{p,q}$ with a fiducial cosmology. Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation considering a constant covariance matrix ($\theta$ stands for the set of cosmological parameters). 
%\begin{minted}[fontsize=\footnotesize]{python}
\begin{lstlisting}[language=iPython]
@jax.jit
def likelihood(p):
    # Create a new cosmology at these parameters
    cosmo = jc.Planck15(Omega_c=p[0], sigma8=p[1])
    # Compute mean and covariance of angular Cls
    mu, C = jc.angular_cl.gaussian_cl_covariance_and_mean(cosmo, ell, tracers, sparse=True)
    # Return likelihood value assuming constant covariance, so we stop the gradient
    # at the level of the precision matrix, and we will not include the logdet term
    # in the likelihood
    P = jc.sparse.inv(jax.lax.stop_gradient(C))
    r = data - mu
    return -0.5 * r.T @ jc.sparse.sparse_dot_vec(P, r)
\end{lstlisting}
%\end{minted}
The \texttt{jc.sparse} functions are implementation of block matrix computations: sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.  The \texttt{jax.jit} decorator is there to build a compiled version of the function at the first use. The goal is to obtain the 1-sigma contour forecast of the two parameters ($\Omega_c, \sigma_8$). To do so, we use the Fisher matrix computation. One way is to use the Hessian of the log-likelihood as follows
\begin{equation}
F_{i,j} = - \frac{\partial^2\mathcal{L}(\theta)}{\partial \theta_i \partial \theta_j}
\qquad (\theta_1=\Omega_c, \theta_2=\sigma_8)
\label{eq:fisher_way1}
\end{equation}
which is mostly  1-line JAX code
\begin{lstlisting}[language=iPython]
hessian_loglik = jax.jit(jax.hessian(likelihood))
F = - hessian_loglik(params)
\end{lstlisting}
An other way uses the computation of the Fisher matrix when the likelihood is a Gaussian distribution as ii is the case above. First, we define a function that computes the mean  $\mu(\ell; \theta)$, and it follows that the Fisher matrix element reads
\begin{equation}
    F_{i,j} = \sum_\ell \frac{\partial \mu(\ell)}{\partial \theta_i}^T C^{-1}(\ell)\frac{\partial \mu(\ell)}{\partial \theta_j}
    \label{eq:fisher_way2}
\end{equation}
where $C^{-1}(\ell)$ is the covariance matrix computed with the fiducial cosmology. In this case one needs also just 1-line of JAX code 
\begin{lstlisting}[language=iPython]
F = jc.sparse.dot(dmu.T, jc.sparse.inv(cov), dmu)
\end{lstlisting}

@@

Once the Fisher matrix can be safely estimated, one can use the MOPED\footnote{Massively Optimised Parameter Estimation and Data compres-
sion} algorithm \citep{2000MNRAS.317..965H,2016PhRvD..93h3525Z, 2017MNRAS.472.4244H}. In the case of the constant covariance matrix the algorithm leads to a  lossless compression which reduces possibly large data vector $x$ (eg. the $N=500$ $C_\ell^{p,q}$ values in the previous section) to $M$ numbers, where $M$ is the number of parameters $\theta_\alpha$ considered (eg. $M=2$ in the previous section). The algorithm computes by iteration $M$ orthogonal vectors of size $N$ such that (taking the notation of the previous section)
\begin{equation}
    b_i = \frac{C^{-1}\mu_{,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)b_j}{\sqrt{F_{i,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)^2}}
\end{equation}
where $\mu_{,i}=\partial \mu/\partial \theta_i$ ($i=1,\dots,M$). The vectors $b_i$, $b_j$ satisfy the following orthogonality constraint
\begin{equation}
    b_i^T C b_j = \delta_{i,j}
\end{equation}
and in turn from the data vector $x=(C_\ell^{p,q})_{p,q,\ell}$ one defines the compressed data set composed of $M$ numbers $(y_i)_{i\leq M}$ according to
\begin{equation}
    y_i = b_i^T x
\end{equation}
These numbers are uncorrelated and of unit variance and $(y_i)_{i\leq M}$ is a set of sufficient statistic for the set of parameters $(\theta_i)_{i\leq M}$. The case where the covariant matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H}
%
\section{Application: Classical Hamiltonian Monte Carlo}
\label{sec:chmc}
%
In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo. We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI). All  the methods have been implemented using the \numpyro\ probabilistic programming language (PPL). 
%
%
%
\subsection{Vanilla HMC}
%
% Show Joe's vanilla HMC results against Cobaya
Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
and developed extensively since.  See \citet{betancourt} for a full review; we describe
very basic features here.
